Isomorphisms as Groupoids: \\ Homotopy Type Theory and Fractional Types

### 2013-10-19 f.tex
f.tex

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage[sans]{dsfont}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}

@@

\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newtheorem{definition}{Definition}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\newcommand{\paths}[1]{\mathds{P}(#1)}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

\begin{document}
\begin{titlepage}
\begin{center}
\vfill
{\LARGE Isomorphisms as Groupoids: \\ 
  Homotopy Type Theory with Fractional Types} \\[1.5cm]
{\Large Jacques Carette${}^*$ \qquad Amr Sabry${}^{\dagger}$}\\[0.5cm]
{\Large $({}^*)$~McMaster University \qquad $({}^\dagger)$~Indiana University}
\vfill
\end{center}
\vfill
\noindent\textbf{Abstract.}

@@

\title{Isomorphisms as Groupoids: \\ 
  Homotopy Type Theory with Fractional Types}
\author{Jacques Carette \and Amr Sabry}
\date{\today}
\maketitle

@@

@@

>>>>

\documentclass[11pt]{article}
\usepackage{fullpage}
\usepackage{url}
\usepackage{stmaryrd}
\usepackage{amssymb}
\usepackage[sans]{dsfont}
\usepackage{color, colortbl}
\usepackage{courier}
\usepackage{graphicx}
\usepackage[all]{xy}
\xyoption{line}

@@

% ----------------------------------------------------------------------
% XY-Pic definitions for making ``wire charts''

\def\wirechart#1#2{\let\labelstyle\objectstyle\xy*[*1.0]\xybox{\small%
\xymatrix@C=10mm@R=4mm#1{#2}}\endxy}
\def\wire#1#2{\ar@{-}[#1]^<>(.5){#2}}
\def\wireright#1#2{\wire{#1}{#2}\ar@{}[#1]|<>(.5)/.6ex/{\dir2{>}}}
\def\wireleft#1#2{\wire{#1}{#2}\ar@{}[#1]|<>(.5)/-.6ex/{\dir2{<}}}
\def\wwblank#1{*=<#1,0mm>{~}}
\def\wblank{\wwblank{11mm}}
\def\blank{\wwblank{8mm}}
\def\vublank{*=<0mm,2.3mm>!D{}}
\def\vdblank{*=<0mm,2.3mm>!U{}}
\def\vsblank{*=<0mm,5.6mm>{}}
\def\wirecross#1{\save[]!L;[#1]!R**@{-}\restore}
\def\wirebraid#1#2{\save[]!L;[#1]!R**{}?(#2)**@{-}\restore\save[#1]!R;[]!L**{}?(#2)**@{-}\restore}
\def\wireopen#1{\save[]!R;[#1]!R**\crv{[]!C&[#1]!C}\restore}
\def\wireclose#1{\save[]!L;[#1]!L**\crv{[]!C&[#1]!C}\restore}
\def\wireopenlabel#1#2{\save[]!R;[#1]!R**\crv{[]!C&[#1]!C}?<>(.5)*^+!R{#2}\restore}
\def\wirecloselabel#1#2{\save[]!L;[#1]!L**\crv{[]!C&[#1]!C}?<>(.5)*^+!L{#2}\restore}
\def\wireid{\blank\save[]!L*{};[]!R*{}**@{-}\restore}
\def\wwireid{\wblank\save[]!L*{};[]!R*{}**@{-}\restore}
\def\wwwireid#1{\wwblank{#1}\save[]!L*{};[]!R*{}**@{-}\restore}
\newcommand{\corner}{\rule{2.5mm}{2.5mm}}
\def\minheight{8mm}
\def\addheight{8mm}
\def\cornerbox#1#2#3{  % draw a box with a marked corner around
                   % the object #1. #2=label
                   % usage:   \ulbox{[ll].[].[ul]}{f}
  \save#1="box";
  "box"!C*+<0mm,\addheight>+=<0mm,\minheight>[|(5)]\frm{-}="box1";
  "box1"*{#2};
  ={"box1"!LU*!LU[]{\corner}}"ul";
  ={"box1"!RU*!RU[]{\corner}}"ur";
  ={"box1"!LD*!LD[]{\corner}}"dl";
  ={"box1"!RD*!RD[]{\corner}}"dr";
  #3
  \restore
}
\def\nnbox#1#2{\cornerbox{#1}{#2}{}}
\def\ulbox#1#2{\cornerbox{#1}{#2}{"ul"}}
\def\urbox#1#2{\cornerbox{#1}{#2}{"ur"}}
\def\dlbox#1#2{\cornerbox{#1}{#2}{"dl"}}
\def\drbox#1#2{\cornerbox{#1}{#2}{"dr"}}
\def\ubox#1#2{\cornerbox{#1}{#2}{"ul","ur"}}
\def\dbox#1#2{\cornerbox{#1}{#2}{"dl","dr"}}
\def\lbox#1#2{\cornerbox{#1}{#2}{"ul","dl"}}
\def\rbox#1#2{\cornerbox{#1}{#2}{"ur","dr"}}
\def\circbox#1{*+[o][F-]{#1}}

\newtheorem{definition}{Definition}
\newtheorem{n-lemma}{Lemma}[section]
\newtheorem{n-example}[n-lemma]{Example}
\newenvironment{example}{\begin{n-example}}{\end{n-example}}

\newcommand{\ssep}{\hspace{1em}} 
\newcommand{\sym}{c}       % braid: could be \sigma or c or \gamma
\newcommand{\cp}{\circ}                 % composition
\newcommand{\symi}{\sym\inv}
\newcommand{\inv}{^{-1}}
\newcommand{\id}{\textrm{\rm id}}       % identity
\newcommand{\x}{\otimes}
\newcommand{\boolt}{\mathit{bool}}
\newcommand{\ffv}{\mathit{false}}
\newcommand{\ttv}{\mathit{true}}
\newcommand{\notb}{\mathit{not}}
\newcommand{\union}{\cup}
\newcommand{\transport}[2]{\overline{#1}(#2)}
\newcommand{\refl}{\mathit{refl}}
\renewcommand{\path}{\leadsto}
\newcommand{\paths}[1]{\mathds{P}(#1)}
%% \renewcommand{\circ}{\circlearrowleft}
\newcommand{\alt}{~|~}
\newcommand{\dt}[1]{\llbracket #1 \rrbracket}
\newcommand{\leftv}[1]{\textsf{left}~#1}
\newcommand{\rightv}[1]{\textsf{right}~#1}
\newcommand{\iso}{\leftrightarrow}
\newcommand{\identlp}{\mathit{identl}_+}
\newcommand{\identrp}{\mathit{identr}_+}
\newcommand{\swapp}{\mathit{swap}_+}
\newcommand{\assoclp}{\mathit{assocl}_+}
\newcommand{\assocrp}{\mathit{assocr}_+}
\newcommand{\identlt}{\mathit{identl}_*}
\newcommand{\identrt}{\mathit{identr}_*}
\newcommand{\swapt}{\mathit{swap}_*}
\newcommand{\assoclt}{\mathit{assocl}_*}
\newcommand{\assocrt}{\mathit{assocr}_*}
\newcommand{\distz}{\mathit{dist}_0}
\newcommand{\factorz}{\mathit{factor}_0}
\newcommand{\dist}{\mathit{dist}}
\newcommand{\factor}{\mathit{factor}}
\newcommand{\idc}{\mathit{id}}
\newcommand{\wedgesum}{\vee}
\newcommand{\smashproduct}{\wedge}
\newcommand{\symc}[1]{\mathit{sym}~#1}
\newcommand{\proves}{\vdash}
\newcommand{\Rule}[4]{
\makebox{{\rm #1}
$\displaystyle
\frac{\begin{array}{l}#2\\\end{array}}
{\begin{array}{l}#3\\\end{array}}$
 #4}}
\newcommand{\jdg}[3]{#2 \proves_{#1} #3}

\begin{document}
\begin{titlepage}
\begin{center}
\vfill
{\LARGE Isomorphisms as Groupoids: \\ 
  Homotopy Type Theory and Fractional Types} \\[1.5cm]
{\Large Jacques Carette${}^*$ \qquad Amr Sabry${}^{\dagger}$}\\[0.5cm]
{\Large $({}^*)$~McMaster University \qquad $({}^\dagger)$~Indiana University}
\vfill
\end{center}
\vfill
\noindent\textbf{Abstract.}

@@

\title{Isomorphisms as Groupoids: \\ 
  Homotopy Type Theory and Fractional Types}
\author{Jacques Carette \and Amr Sabry}
\date{\today}
\maketitle

@@

From the wikipedia page on homeomorphism: 
\begin{quote}
The intuitive criterion of stretching, bending, cutting and gluing back
together takes a certain amount of practice to apply correctly—it may not be
obvious from the description above that deforming a line segment to a point
is impermissible, for instance. It is thus important to realize that it is
the formal definition given above that counts.

This characterization of a homeomorphism often leads to confusion with the
concept of homotopy, which is actually defined as a continuous deformation,
but from one function to another, rather than one space to another. In the
case of a homeomorphism, envisioning a continuous deformation is a mental
tool for keeping track of which points on space X correspond to which points
on Y—one just follows them as X deforms. In the case of homotopy, the
continuous deformation from one map to the other is of the essence, and it is
also less restrictive, since none of the maps involved need to be one-to-one
or onto. Homotopy does lead to a relation on spaces: homotopy equivalence.

There is a name for the kind of deformation involved in visualizing a
homeomorphism. It is (except when cutting and regluing are required) an
isotopy between the identity map on X and the homeomorphism from X to Y.
\end{quote}

It is strange for homotopy type theory to be all about isomorphisms and then
not insist on isomorphisms between spaces.

If we focus on homeomorphisms, the higher-order space is a \emph{torsor}.

@@

%% the next 7 figures are taken from Selinger's paper on graphical languages
%% for monoidal categories; if we use them make sure it's ok with him and
%% acknowledge him

In homotopy type theory, there is a focus on isomomorphisms represented by
paths (proofs of identities) and yet arbitrary functions are considered. We
propose to only consider functions that are themselves isomorphisms: this
will allow us to view the function itself as a collection of paths and it
doesn't lose anything in terms of computability because we know that we could
do everything with reversible functions if we wanted to.

Consider the function $\notb$ from the space $\boolt$ to itself. We can
represent this function as a pair of paths. Since we are in a topological
world, we should care whether the path top pass crosses over the bottom one
or vice-versa. In the diagram below, we chose for the top wire to cross over
the bottom one:

\begin{center}
  \begin{tabular}{@{}llc@{}}
    $\notb : \boolt \rightarrow \boolt$ & & 
    $\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
        *{}\wireright{r}{\ffv}&\blank\wirecross{d}\wireright{r}{\ffv}&\\
        *{}\wireright{r}{\ttv}&\blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
        }}$ 
  \end{tabular}
\end{center}

It is important also to note that the two paths must be taken together
because there is a constraint between them. If the top path takes $\ffv$ to
$\ttv$ then the bottom path cannot take $\ttv$ to $\ttv$: it \emph{has} to
lead to $\ffv$. We will return to this point later.

For now, consider the functions:
\[\begin{array}{rcl}
\mathit{id}^1 &=& \lambda x.x \\
\mathit{id}^2 &=& \lambda x. \notb~(\notb~x)
\end{array}\]
These two functions are evidently extensionally equivalent and hence in the
conventional treatment of homotopy type theory, it is acceptable to assume
they are also propositionally equivalent, i.e., it is acceptable to postulate
a path $\mathit{id}^1 \path \mathit{id}^2$ in the space $\boolt \rightarrow
\boolt$.

When represented as paths, we get the following situation for the second
function:
\[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
    *{}\wireright{r}{\ffv}&
    \blank\wirecross{d}\wireright{r}{\ttv}&
    \blank\wirecross{d}\wireright{r}{\ffv}&
    \\
    *{}\wireright{r}{\ttv}&
    \blank\wirebraid{u}{.3}\wireright{r}{\ffv}&
    \blank\wirebraid{u}{.3}\wireright{r}{\ttv}&
    \\
    }}
\]
Repeating the $\notb$ operation results in a twisted path that externally
behaves like the identity but that evidently not identical to the straight
path with no twists. Thus although one can justify that the two functions
should have a path between them, the function extensionality axiom does not
give insights into \emph{what} this path is actually doing topologically.

%% \begin{center}
%%   \begin{tabular}{@{}llc@{}}
%%     Braiding & $\sym_{A,B}$ &
%%     $\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
%%         *{}\wireright{r}{B}&\blank\wirecross{d}\wireright{r}{A}&\\
%%         *{}\wireright{r}{A}&\blank\wirebraid{u}{.3}\wireright{r}{B}&
%%         }}$ \\
%%   \end{tabular}
%% \end{center}

%% Note that the braiding satisfies
%% $\sym_{A,B}\cp\symi_{A,B}=\id_{A\x B}$, but not
%% $\sym_{A,B}\cp\sym_{B,A}=\id_{A\x B}$. Graphically:

%% \[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
%%     *{}\wireright{r}{B}&
%%     \blank\wirecross{d}\wireright{r}{A}&
%%     \blank\wirebraid{d}{.3}\wireright{r}{B}&
%%     \\
%%     *{}\wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
%%     \blank\wirecross{u}\wireright{r}{A}&
%%     \\
%%     }}
%% = \id_{A\x B},
%% \]
%% \[\vcenter{\wirechart{@C=1.5cm@R=0.8cm}{
%%     *{}\wireright{r}{B}&
%%     \blank\wirecross{d}\wireright{r}{A}&
%%     \blank\wirecross{d}\wireright{r}{B}&
%%     \\
%%     *{}\wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{A}&
%%     \\
%%     }}
%% \neq \id_{A\x B}.
%% \]

%% \begin{example}
%%   The hexagon axiom translates into the following in the graphical
%%   language:

%%   \[ (\id_B\x\sym_{A,C}) \cp \alpha_{B,A,C} \cp (\sym_{A,B}\x\id_C)
%%   = \alpha_{B,C,A} \cp (\sym_{B,C\x A}) \cp \alpha_{A,B,C}
%%   \]
%%   \[\vcenter{\wirechart{@R=6mm}{
%%     \wireright{r}{C}&
%%     \wireid\wireright{r}{C}&
%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
%%     \\
%%     \wireright{r}{B}&
%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
%%     \\
%%     \wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
%%     \wireid\wireright{r}{B}&
%%     \\
%%     }}
%%   \ssep=\ssep
%%   \vcenter{\wirechart{@R=6mm}{
%%     \wireright{r}{C}&
%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
%%     \\
%%     \wireright{r}{B}&
%%     \blank\wirecross{d}{.3}\wireright{r}{C}&
%%     \\
%%     \wireright{r}{A}&
%%     \blank\wirebraid{uu}{.2}\wireright{r}{B}&
%%     \\
%%     }}
%%   \]
%% \end{example}

%% \begin{example}
%%   The {\em Yang-Baxter equation} is the following equation, which is a
%%   consequence of the hexagon axiom and naturality:
%%   \[ (\sym_{B,C}\x\id_A)\cp(\id_B\x\sym_{A,C})\cp(\sym_{A,B}\x\id_C) =
%%   (\id_C\x\sym_{A,B})\cp(\sym_{A,C}\x\id_B)\cp(\id_A\x\sym_{B,C}).
%%   \]
%%   In the graphical language, it becomes:
%%   \[\vcenter{\wirechart{@C-4mm@R=6mm}{
%%     \wireright{r}{C}&
%%     \wireid\wireright{r}{C}&
%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
%%     \wireid\wireright{r}{A}&
%%     \\
%%     \wireright{r}{B}&
%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
%%     \blank\wirecross{d}{.3}\wireright{r}{B}&
%%     \\
%%     \wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
%%     \wireid\wireright{r}{B}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
%%     \\
%%     }}
%%   \ssep=\ssep
%%   \vcenter{\wirechart{@C-4mm@R=6mm}{
%%     \wireright{r}{C}&
%%     \blank\wirecross{d}{.3}\wireright{r}{B}&
%%     \wireid\wireright{r}{B}&
%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
%%     \\
%%     \wireright{r}{B}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
%%     \blank\wirecross{d}{.3}\wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{B}&
%%     \\
%%     \wireright{r}{A}&
%%     \wireid\wireright{r}{A}&
%%     \blank\wirebraid{u}{.3}\wireright{r}{C}&
%%     \wireid\wireright{r}{C}&
%%     \\
%%     }}
%%   \]
%% \end{example}
